---
title: "R Programming"
author: "Furkan Oguz"
date: "4/20/2020"
output: pdf_document
---

# R Porgramming
## Week 1: Getting Started and R Nuts and Bolts
### Writing Code/Setting Your Working Directory

In R (not R Studio): How to set my working directory and how to edit R code files using the text editor.

You can find out what your working directory is by using following code:
```{r}
getwd()
```

To change the working directory: Go to (above) Misc > Change Working Directory. 

Code to list all files within directory:
```{r}
dir()
```

When you want to look if there are objects in your work space:
```{r}
ls()
```

You can open a text editor in R with the white paper sign. To run it you can copy the content and run on R or you can save the file as .r-file. To source the file you saved, you can source it within the directory with the following code:
```{r}
## source("filename.R")
```

***

### Learning Objectives
* Install R and RStudio software packages
* Download and install the swirl package for R
* Describe the history of the S and R programming lectures
* Describe the differences between atomic data types
* Execute basiv arithmetic operations
* Subset R objects using the "[", "[[" and "$" operators and logical vectors
* Describe the explicit coercion feature of R
* Remove missing (NA) values from a vector

***

### Introduction
Course specializiation is to get the nuts and bolts of R programming!

***

### Overview and History of R
**What is R?** It is a dialect of S.
**What is S?** It is a language initiated as an internal statistical analysis environment (originally implemented as Fortran libraries -> Matlab) In earliest version did not contain functions for statistical modelling. Later it was rewritten in C to make it more compatibel. A later version (1998) was that one we use today.

**S Philosohpy:** Would enter language by beginning in an interactive environment. So sliding into programming!

**Back to R:** Invented in 1991. First public announcement 1993. In 1995 it is a free software!

**Features of R:** 
* Syntax is very similar to S
* Semantics are superficially similar to S, but in reality quite different.
* Runs on almost any standard computing platform/OS
* Frequent releases and active development!
* Quite lean, functionality is divided into modular packages
* Graphicscapabilities very sophisitcated and better than most stat packages
* Useful for interactice work, but contains a powerful programming language for developing new tools (user -> programmer)
* Very active and vibrant user community (Stack Overflow)
* It's free


**Free Software:** 
* Freedom to run the program, for any purpose (freedom 0)
* Freedom to study how the program works, and adapt it to your needs (freedom 1). Acces to the source code.
* Freedom to redistribute copies so you can help your neighbour (freedom2)
* Freedom to improve the program, and release your improvements to the public, so that the whole community benefits (freedom 3).

**Drawbacks of R:**
* 40 year old technology
* Little built-in support for dynamic or 3D graphis (but improved greatly)
* Functionality is based on consumer demand and user contributions
* Objects must be stored in physical memory (but great advancements)
* Not ideal for all possible situations (but for all software packages same problem)

**Design of the R System:**
It is divided into 2 conceptual parts:
* "Base* R system that you download from CRAN
* Everything else
R functionality is divided into a number of packages:
* "Base" system contains base package with most fundamental functions
* Other packages contained in the "base" system include **utils, stats, datasets, graphics, grDevices, grid, methods, tools, parallel, compiler, splines, tcltk, stats4**
* Also "recommended" packages: **boot, class, cluster, codetools, foreign, KernSmooth, lattice, mgcv, nlme, rpart, survival, MASS, spatial, nnet, Matrix**
* Many other packages (4'000)

***

### Getting Help
Email: r-help@r-project.org

***

### R Console Input and Evaluation

**Entering Input:**
Things we type in R prompt are called expressions. The sympobl looking like left-hand-arrow is actually the assignment operator.
```{r}
x <- 1
print(x)
msg <- "hello"
z <- ## Incomplete expression
```
This means tha value right is assigned to variable left. The last example is an error, because it is a comment which R doesn't use for computation.

**Evaluation:**
When expression is entered in prompt, it is evaluated and the result of evaluated expression is returned. Result may be auto-printed.
```{r}
x <- 5 ## nothing printed
x ## auto-printing occurs (typing objects name!)
print(x) ## explicit printing
```

**Printing:**
The : operator is use to create integer sequences. In this case integer vector
```{r}
x <- 1:20
x
```

***

### Data Types - R Objects and Attributes
**Objects:** All the things we that we encounter in R. There are different kind of data, but everything is an object.
R has 5 basic or "atomic" classes of objects:
* character
* numeric (real numbers)
* integer
* complex
* logical (True/False)

The most basic object is a vector:
* Can only obtain objects of the same class
* BUT: the one exception is a *list*, which is represented as vector, but can contain objects of different classes (reason why usually used, you can even have a list within a list, very useful!)

How to create empty vector by using vector function:
```{r}
vector()
```

**Numbers:**
3 Numbers are treated as numeric objects (double precision real numbers)
* If explicitly integers were wished, you need to specify the L suffix.
* Example: Entering 1 gives numeric object; entering 1L gives integer
* There is also a special number: Inf which is infinity like 1/0. It can be used in ordinary calculations like 1/Inf which is 0.
* The value NaN is undefined value ("not a number"), 0/0 is NaN. Can also be thaught as  a missing value.

**Attributes:**
Objects can have attributes
* names, dimnames
* dimensions ( matrices, arrays so on)
* class
* length
* other user-defined attributes/metadata

Attributes of an object can be acceses with **attributes() function**!

***

### Data Types - Vectors and Lists
**Creating Vectors:**
The **c() function** can also be used to create vectors and objects
```{r}
x <- c(0.5, 0.6) ##numeric
x <- c(TRUE, FALSE) ##logical
x <- c(T, F) ##logical
x <- c("a", "b", "c") ##character
x <- 9:29 ##integer
x <- c(1+0i, 2+4i) ##complex
```

Using the **vector()** function:
```{r}
x <- vector("numeric", length=20)
x
```

**Mixing Objects:**
```{r}
y <- c(1.7, "a") ##character
y <- c(TRUE, 2) ##numeric
y <- c("a", TRUE) ##character
```

When different objects are mixed in a vector, *coercion* occurs so that every element in the vector is of the same class!! (TRUE is a 1, and FALSE is a 0, they will be converted as a number)

**Explicit Coercion:**
Objects can be explicitly coerced from one class to another using the as.*functions, if available.

```{r}
x<- 0:6
class(x)
as.numeric(x)
as.logical(x)
as.character(x)
```

Nonsensical coercion results in NAS.

```{r}
x <- c("a", "b", "c")
as.numeric(x)
as.logical(x)
as.complex(x)
```

**Lists:**
Lists are a special type of vector that can contain elements of different classes. Lists are a very important data type in R!!

```{r}
x <-list(1, "a", TRUE, 1+4i)
x
```
Lists have for index, always double brackets on them!

***

### Data Types - Matrices

Matrices are vectors with a dimension attribute (itself an integer vector of length 2, **nrow** and **ncol**).

```{r}
m <- matrix(nrow=2, ncol=3)
m
dim(m)
attributes(m)
```

Matrices are constructed *column-wise*, so entries can be thought of starting in the "upper left" corner and running down the columns.

```{r}
m <- matrix(1:6, nrow=2, ncol=3)
m
```

It can also be created directly from vectors by adding a dimension attribute.

```{r}
m <- 1:10
m
dim(m) <- c(2,5)
m
```

Can be created by *column-binding* or *row-binding* with **cbind()** and **rbind()**

```{r}
x <- 1:3
y <- 10:12
cbind(x,y)
rbind(x,y)
```

***

### Data Types -Factors

Factors are used to represent categorical data. Can be unordered or ordered. One can think of a factor as an integer vector where each integer has a *label*.
* Factors are treated specially by modelling functions **lm()** and **glm()**
* Using factors with labels is *better* than using integers because factors are self-decribing; having a variable that has values "Male" and "Female" is better than a variable that has values 1 and 2.

```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"))
x
table(x)
unclass(x) ##brings an integer vector
```

The order of the levels can be set using the **levels** argument to **factor()**. Can be important in linear modelling because the first level is used as the baseline level.

```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"), levels=c("yes", "no"))
x
```

***

### Data Types - Missing Values

**Missing Values:**
Are denoted by **NA** or **NaN** for undefined mathematical operations.
* **is.na()** is used to test objects if they are **NA**
* **is.nan()** is used to test for **NaN**
* **NA** values have a class also, so there are integer **NA**, character **NA** etc.
* A **NaN** value is also a **NA** but the converse is not true

```{r}
x <- c(1, 2, NA, 10, 3)
is.na(x)
is.nan(x)
x <- c(1, 2, NaN, NA, 4)
is.na(x)
is.nan(x)
```

***

### Data Types - Data Frames

Data frames are used to store tabular data
* They are represented as a special type of list where every element of the list has to have the same length
* Each element of the list can be thought of as a column and the length of each element of the list is the number of rows
* Unlike matrices, data frames can store different class of objects in each column (like lists); matrices must have every element be the same class
* Data frames also have a special attribute called **row.names**
* Are usually created by calling **read.table(*)** or **read.csv()**
* Can be converted to a matrix by calling **data.matrix()**

```{r}
x <- data.frame(foo=1:4, bar=c(T, T, F, F))
x
nrow(x)
ncol(x)
```

***

### Data Types - Names Attribute

R objects can also have names, which is very useful for writing readable code and self-describing objects.

```{r}
x <- 1:3
names(x)
names(x) <- c("foo", "bar", "norf")
x
names(x)
```

Lists can also have names.

```{r}
x <- list(a=1, b=2, c=3)
x
```

And matrices.

```{r}
m <- matrix(1:4, nrow=2, ncol=2)
dimnames(m) <- list(c("a", "b"), c("c", "d"))
m
```

***

### Data Types - Summary

* atomic classes: numeric, logical, character, integer, complex
* vectors, lists
* factors
* missing values
* data frames
* names

***

### Reading Tabular Data

There are a few principal functions reading data into R:
* **read.table**, **read.csv** for reading tabular data
* **readLines**, for reading lines of a text file
* **source**, for reading in R code files (*inverse of dump*)
* **dget**, for reading R code files (*inverse of dput*) (it's for reading R objects that have been dparsed into text files)
* **load**, for reading in saved workspaces
* **unserialize**, for reading single R objects in binary form

There are analogous functions for writing data to files:
* **write.table**
* **writeLines**
* **dump**
* **dput**
* **save**
* **serialize**

Reading Data Files with *read.table* (very important!!):
It is the most commonly used functions for reading data. It has a few important arguments:
* **file**, the name of a file, or a connection
* **header**, logical indicating if the file has a header line
* **sep**, a string indicating how the columns are separated
* **colClasses**, a character vector indicating the class of each column in the dataset
* **nrows**, the number of rows in the dataset
* **comment.char**, a character string indicating the comment character
* **skip**, the number of lines to skip from the beginning
* **strinsAsFactors**, should character variables be coded as factors?



